---
title: "A Pattern Based Assessment of Car Crashes in Centre County (2023-2024)"
author: "Akil Creswell, Jooan Choi, Inayat Roy"
format: pdf
editor: visual
---

## Overview

This project will set to answer the questions: "What patterns in collision type, time of day, and road conditions can explain crashes of varying severity in Centre County from 2023 to 2024?"

### **Interesting Insight**

::: {.callout-note appearance="minimal" icon="false"}
Drivers hitting a fixed object(non moving) was the most common collision type on every day of the week. This specific collision type happened on mostly Fridays and Saturdays
:::

Crashes that were categorized as "Hit Fixed Object" accounted for 722 total incidents in 2023-2024. This was the highest collision count(by type) of the ten in the datasets. Friday and Saturday combined for 225 of those crashes, showing a end-of-week behavioral pattern.

//insight-chart
```{r}
#| label: insight-chart
#| echo: false
#| message: false
#| warning: false
#| fig-alt: "Line chart showing the frequency of each collision type across days of the week in Centre County, Pennsylvania for 2023 and 2024. Hit Fixed Object is the highest line on every day, peaking on Friday and Saturday. Rear-end and Angle collisions are the next most frequent. All other collision types remain low and relatively flat across the week."
# Style Guide: Google R Style Guide
# Primary Author: Akil Creswell
# Reviewed by: Jooan Choi

library(ggplot2)
library(dplyr)
library(readr)

collision_labels <- c(
  "0" = "Non-collision",
  "1" = "Rear-end",
  "2" = "Head-on",
  "3" = "Backing",
  "4" = "Angle",
  "5" = "Sideswipe",
  "6" = "Sideswipe (Opposite)",
  "7" = "Hit Fixed Object",
  "98" = "Other",
  "8" = "Hit Non-Motorist"
)

day_labels <- c(
  "1" = "Sun",
  "2" = "Mon",
  "3" = "Tue",
  "4" = "Wed",
  "5" = "Thu",
  "6" = "Fri",
  "7" = "Sat"
)

crashRaw2023 <- read_csv("CRASH_CENTRE_2023.csv")
crashRaw2024 <- read_csv("CRASH_CENTRE_2024.csv")
crashes <- bind_rows(crash_2023, crash_2024)

crash_summary <- crashes %>%
  mutate(
    collision_label = recode(
      as.character(COLLISION_TYPE),
      !!!collision_labels,
      .default = "Other"
    ),
    day_label = factor(
      recode(as.character(DAY_OF_WEEK), !!!day_labels),
      levels = c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat")
    )
  ) %>%
  filter(!is.na(day_label), !is.na(collision_label)) %>%
  group_by(day_label, collision_label) %>%
  summarise(count = n(), .groups = "drop")

ggplot(crash_summary, aes(
  x = day_label,
  y = count,
  color = collision_label,
  group = collision_label
)) +
  geom_line(linewidth = 0.5, na.rm = TRUE) +
  geom_point(aes(shape = collision_label), size = 1, na.rm = TRUE) +
  labs(
    title = "Collision Type Frequency by Day of Week",
    subtitle = "Centre County, Pennsylvania | 2023-2024",
    x = "Day of Week",
    y = "Number of Crashes",
    color = "Collision Type",
    caption = "Source: PennDOT Crash Data",
    shape = "Collision Type"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 8),
    axis.text = element_text(size = 10)
  )

//data-wrangling
```{r}
# Centre County Crash Data Wrangling ----
## Import and clean 2024 Centre County crash data
## for crash severity and collision type analysis
# Style Guide: Google R Style Guide
# Primary Author: Jooan Choi 
# Reviewed by: Akil Creswell

rm(list = ls())

# Step 1: Load Packages ----
## Needed Packages: tidyverse
library(tidyverse)

# Step 2: Load Data ----
## Needed data: 2024 Centre County crash data

#setwd("C:/Users/jooan/OneDrive - The Pennsylvania State University/SPRING 26/STAT 184/project")
crashRaw2023 <- read_csv("CRASH_CENTRE_2023.csv")
crashRaw2024 <- read_csv("CRASH_CENTRE_2024.csv")

# Combine datasets
crashRaw <- bind_rows(
  crashRaw2023,
  crashRaw2024
)
#view(crashRaw)

# Step 3: Make code tables ----
## These tables translate coded values into readable labels

month <- data.frame(
  CRASH_MONTH = 1:12,
  Month = c(
    "January", "February", "March", "April",
    "May", "June", "July", "August",
    "September", "October", "November", "December"
  )
)

day <- data.frame(
  DAY_OF_WEEK = 1:7,
  Day = c(
    "Sunday", "Monday", "Tuesday", "Wednesday",
    "Thursday", "Friday", "Saturday"
  )
)

collision <- data.frame(
  COLLISION_TYPE = c(0, 1, 2, 3, 4, 5, 6, 7, 8, 98),
  Collision_Type = c(
    "Non-collision",
    "Rear-end",
    "Head-on",
    "Backing",
    "Angle",
    "Sideswipe same direction",
    "Sideswipe opposite direction",
    "Hit fixed object",
    "Hit non-motorist",
    "Other"
  )
)

severity <- data.frame(
  MAX_SEVERITY_LEVEL = c(0, 1, 2, 3, 4, 8, 9),
  Severity_Level = c(
    "Property damage only",
    "Fatal",
    "Suspected serious injury",
    "Suspected minor injury",
    "Possible injury",
    "Injury unknown severity",
    "Unknown if injured"
  )
)

weather <- data.frame(
  WEATHER1 = c(2, 3, 4, 5, 6, 7, 8, 9, 10, 98, 99),
  Weather = c(
    "Blowing snow",
    "Clear",
    "Cloudy",
    "Fog, smog, or smoke",
    "Freezing rain or sleet",
    "Rain",
    "Severe crosswinds",
    "Sleet or hail",
    "Snow",
    "Other",
    "Unknown"
  )
)

road <- data.frame(
  ROAD_CONDITION = c(1, 2, 3, 6, 7, 8, 9, 98, 99),
  Road_Condition = c(
    "Dry",
    "Ice or frost",
    "Mud, dirt, or gravel",
    "Slush",
    "Snow",
    "Water",
    "Wet",
    "Other",
    "Unknown"
  )
)

area <- data.frame(
  URBAN_RURAL = c(1, 2, 3),
  Area_Type = c(
    "Rural",
    "Urbanized",
    "Urban"
  )
)

# Step 4: Select variables ----
crashProgress <- crashRaw |>
  select(
    CRN,
    CRASH_YEAR,
    CRASH_MONTH,
    DAY_OF_WEEK,
    COLLISION_TYPE,
    MAX_SEVERITY_LEVEL,
    INJURY_COUNT,
    FATAL_COUNT,
    WEATHER1,
    ROAD_CONDITION,
    URBAN_RURAL,
    VEHICLE_COUNT,
    TOTAL_UNITS
  )

# Step 5: Convert variables ----
crashProgress2 <- crashProgress |>
  mutate(
    CRN = as.character(CRN),
    CRASH_YEAR = as.numeric(CRASH_YEAR),
    CRASH_MONTH = as.numeric(CRASH_MONTH),
    DAY_OF_WEEK = as.numeric(DAY_OF_WEEK),
    COLLISION_TYPE = as.numeric(COLLISION_TYPE),
    MAX_SEVERITY_LEVEL = as.numeric(MAX_SEVERITY_LEVEL),
    INJURY_COUNT = as.numeric(INJURY_COUNT),
    FATAL_COUNT = as.numeric(FATAL_COUNT),
    WEATHER1 = as.numeric(WEATHER1),
    ROAD_CONDITION = as.numeric(ROAD_CONDITION),
    URBAN_RURAL = as.numeric(URBAN_RURAL),
    VEHICLE_COUNT = as.numeric(VEHICLE_COUNT),
    TOTAL_UNITS = as.numeric(TOTAL_UNITS)
  ) |>
  filter(
    (CRASH_YEAR == 2023 | CRASH_YEAR == 2024),
    !is.na(CRN)
  )

# Step 6: Join code tables ----
crashProgress3 <- crashProgress2 |>
  left_join(month, by = "CRASH_MONTH") |>
  left_join(day, by = "DAY_OF_WEEK") |>
  left_join(collision, by = "COLLISION_TYPE") |>
  left_join(severity, by = "MAX_SEVERITY_LEVEL") |>
  left_join(weather, by = "WEATHER1") |>
  left_join(road, by = "ROAD_CONDITION") |>
  left_join(area, by = "URBAN_RURAL")

# Step 7: Severity group ----
crashProgress4 <- crashProgress3 |>
  mutate(
    Severity_Group = if_else(
      condition = MAX_SEVERITY_LEVEL == 1 | MAX_SEVERITY_LEVEL == 2,
      true = "More severe",
      false = "Less severe"
    )
  ) |>
  mutate(
    Severity_Group = if_else(
      condition = MAX_SEVERITY_LEVEL == 8 | MAX_SEVERITY_LEVEL == 9,
      true = "Unknown",
      false = Severity_Group
    )
  )

# Step 8: Final dataset ---
crashClean <- crashProgress4 |>
  mutate(
    Year = CRASH_YEAR
  ) |>
  select(
    CRN,
    Year,
    Month,
    Day,
    Collision_Type,
    Severity_Level,
    Severity_Group,
    INJURY_COUNT,
    FATAL_COUNT,
    Weather,
    Road_Condition,
    Area_Type,
    VEHICLE_COUNT,
    TOTAL_UNITS
  )
```
