---
title: "A Pattern Based Assessment of Car Crashes in Centre County (2023-2024)"
author: "Akil Creswell, Jooan Choi, Inayat Roy"
format:
  pdf:
    pdf-engine: xelatex
    mainfont: "Times New Roman"
    fontsize: 12pt
    include-in-header:
      text: |
        \usepackage{xcolor}
        \usepackage{caption}
        \usepackage{colortbl}
        \usepackage{booktabs}
        \usepackage{longtable}
        \usepackage{array}
        \usepackage{multirow}
        \usepackage{float}
        \usepackage{sectsty}
        \subsectionfont{\centering\LARGE}
        \subsubsectionfont{\centering\large}

editor: visual
bibliography: J_A_I_STAT184Project_Datasets/references.bib
csl: J_A_I_STAT184Project_Datasets/formatting/MLA9.csl
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false
# Style Guide: Google R Style Guide
# Primary Author: Akil Creswell
# Reviewed by: Jooan Choi, Inayat Roy

library(tidyverse)
library(knitr)
library(kableExtra)
library(ggplot2)
library(dplyr)
library(readr)
```

## Introduction {#sec-introduction}

From 2023-2024 there was a recorded 2,097 crashes, which range from minor property damage to fatal collisions. The main goal of this project is examine crash records collected by Pennsylvania Department of Transportation (PennDOT) to answer the following research question: **What patterns in collision type and road conditions characterize crashes of varying severity in Centre County from 2023 to 2024?**

The secondary goal is to explore the data to be able discern possible ways to curb accidents in Centre County. The conditions are seemingly never consistent due to unpredictable weather patterns in the area, which makes condition based analysis like this one important.

This analysis follows the Exploratory Data Analysis paradigm and uses the Tidyverse coding paradigm in R, following the Google R Style Guide.

{{< pagebreak >}}

## Data

### Data Provenance

The data used in this project was obtained for the Pennsylvania Department of Transportation (PennDOT) crash data portal [@penndot2026]. Two CSV files were downloaded: `CRASH_CENTRE_2023.csv`, containing 1,025 crash records, and `CRASH_CENTRE_2024.csv`, containing 1,072 crash records, to combine for a total of 2,097 records across both years. Each record represents a unique crash event with ID, Crash Record Number (`CRN`).

Local Centre County law enforcement collected the data and submitted it into PennDOT's statewide systems. To interpret the data we used the data dictionary provided by PennDOT, which was obtained from the same portal.

### FAIR and CARE Principles

**FAIR**

There are several ways the PennDOT data satisfies FAIR principles. The data are *findable* and *accessible* with the portal being publicly accessible, and organized (county, year). The data are *interpretable* in CSV format and a given data dictionary to make data simple to read. Reproducibility is supported given that CSVs can be downloaded from one URL.

A challenge the principle of *reusability* is that multiple columns are populated with mostly null values making variations in scope difficult.

**CARE**

This dataset does not directly involve Indigenous communities as there is no information about the people involved in the car crashes. It focuses purely on the confounding factors of the crashes. Despite that, the conclusions from this analysis should not be used to perpetuate any negativity of any kind towards any people, groups, or communities.

{{< pagebreak >}}

## Case and Attribute Description

Each case in this dataset is an individual car crash recorded in Centre County, Pennsylvania between January 1st, 2023 - December 31st, 2024 identified by Crash Record Number(`CRN`). Each unit is a crash, not a vehicle nor individual.

The following attributes were chosen to contribute to answering our research question.\

- `Collision_Type`

  - How the crash occurred

  - 10 unique categories

- `Severity_Level`

  - Highest injury level sustained in crash

  - 7 unique categories

- `Severity_Group`

  - Derives from `MAX_SEVERITY_LEVEL`

  - Groups containing injury severity level

    - `More Severe`

      - 1:Fatal

      - 2: Serious

    - `Less Severe`

      - 0: Property damage

      - 3: Minor injury

      - 4: Possible injury

    - `Unknown`

      - 8: Injury unknown severity

      - 9: Unknown

- `Road_Condition`

  - Surface condition of the roadway

  - 9 unique categories

- `Day`

  - Day of the week the crash occurred

- `Month`

  - Month the crash occurred

- `Area_Type`

  - Rural or Urbanized classification of crash location

- `INJURY_COUNT`

  - Total injuries occurred in crash

    - data type: numeric

- `FATAL_COUNT`

  - Total fatalities occurred in crash

    - data type: numeric

{{< pagebreak >}}

```{r}
#| label: data-wrangling
#| message: false
#| warning: false
#| include: false
# Centre County Crash Data Wrangling ----
## Import and clean 2024 Centre County crash data
## for crash severity and collision type analysis
# Style Guide: Google R Style Guide
# Primary Author: Jooan Choi 
# Reviewed by: Akil Creswell


# Step 1: Load Data ----
## Needed data: 2024 Centre County crash data

crashRaw2023 <- read_csv("J_A_I_STAT184Project_Datasets/CRASH_CENTRE_2023.csv")
crashRaw2024 <- read_csv("J_A_I_STAT184Project_Datasets/CRASH_CENTRE_2024.csv")

# Combine datasets
crashRaw <- bind_rows(
  crashRaw2023,
  crashRaw2024
)
glimpse(crashRaw)

# Step 2: Make code tables ----
## These tables translate coded values into readable labels

month <- data.frame(
  CRASH_MONTH = 1:12,
  Month = c(
    "January", "February", "March", "April",
    "May", "June", "July", "August",
    "September", "October", "November", "December"
  )
)

day <- data.frame(
  DAY_OF_WEEK = 1:7,
  Day = c(
    "Sunday", "Monday", "Tuesday", "Wednesday",
    "Thursday", "Friday", "Saturday"
  )
)

collision <- data.frame(
  COLLISION_TYPE = c(0, 1, 2, 3, 4, 5, 6, 7, 8, 98),
  Collision_Type = c(
    "Non-collision",
    "Rear-end",
    "Head-on",
    "Backing",
    "Angle",
    "Sideswipe same direction",
    "Sideswipe opposite direction",
    "Hit fixed object",
    "Hit non-motorist",
    "Other"
  )
)

severity <- data.frame(
  MAX_SEVERITY_LEVEL = c(0, 1, 2, 3, 4, 8, 9),
  Severity_Level = c(
    "Property damage only",
    "Fatal",
    "Suspected serious injury",
    "Suspected minor injury",
    "Possible injury",
    "Injury unknown severity",
    "Unknown if injured"
  )
)

road <- data.frame(
  ROAD_CONDITION = c(1, 2, 3, 6, 7, 8, 9, 98, 99),
  Road_Condition = c(
    "Dry",
    "Ice or frost",
    "Mud, dirt, or gravel",
    "Slush",
    "Snow",
    "Water",
    "Wet",
    "Other",
    "Unknown"
  )
)

area <- data.frame(
  URBAN_RURAL = c(1, 2, 3),
  Area_Type = c(
    "Rural",
    "Urbanized",
    "Urban"
  )
)

# Step 3: Select variables ----
crashProgress <- crashRaw |>
  select(
    CRN,
    CRASH_YEAR,
    CRASH_MONTH,
    DAY_OF_WEEK,
    COLLISION_TYPE,
    MAX_SEVERITY_LEVEL,
    INJURY_COUNT,
    FATAL_COUNT,
    ROAD_CONDITION,
    URBAN_RURAL,
    VEHICLE_COUNT,
    TOTAL_UNITS
  )

# Step 4: Convert variables ----
crashProgress2 <- crashProgress |>
  mutate(
    CRN = as.character(CRN),
    CRASH_YEAR = as.numeric(CRASH_YEAR),
    CRASH_MONTH = as.numeric(CRASH_MONTH),
    DAY_OF_WEEK = as.numeric(DAY_OF_WEEK),
    COLLISION_TYPE = as.numeric(COLLISION_TYPE),
    MAX_SEVERITY_LEVEL = as.numeric(MAX_SEVERITY_LEVEL),
    INJURY_COUNT = as.numeric(INJURY_COUNT),
    FATAL_COUNT = as.numeric(FATAL_COUNT),
    ROAD_CONDITION = as.numeric(ROAD_CONDITION),
    URBAN_RURAL = as.numeric(URBAN_RURAL),
    VEHICLE_COUNT = as.numeric(VEHICLE_COUNT),
    TOTAL_UNITS = as.numeric(TOTAL_UNITS)
  ) |>
  filter(
    (CRASH_YEAR == 2023 | CRASH_YEAR == 2024),
    !is.na(CRN)
  )

# Step 5: Join code tables ----
crashProgress3 <- crashProgress2 |>
  left_join(month, by = "CRASH_MONTH") |>
  left_join(day, by = "DAY_OF_WEEK") |>
  left_join(collision, by = "COLLISION_TYPE") |>
  left_join(severity, by = "MAX_SEVERITY_LEVEL") |>
  left_join(road, by = "ROAD_CONDITION") |>
  left_join(area, by = "URBAN_RURAL")

# Step 6: Severity group ----
crashProgress4 <- crashProgress3 |>
  mutate(
    Severity_Group = if_else(
      condition = MAX_SEVERITY_LEVEL == 1 | MAX_SEVERITY_LEVEL == 2,
      true = "More severe",
      false = "Less severe"
    )
  ) |>
  mutate(
    Severity_Group = if_else(
      condition = MAX_SEVERITY_LEVEL == 8 | MAX_SEVERITY_LEVEL == 9,
      true = "Unknown",
      false = Severity_Group
    )
  )

# Step 7: Final dataset ---
crashClean <- crashProgress4 |>
  mutate(
    Year = CRASH_YEAR
  ) |>
  select(
    CRN,
    Year,
    Month,
    Day,
    Collision_Type,
    Severity_Level,
    Severity_Group,
    INJURY_COUNT,
    FATAL_COUNT,
    Road_Condition,
    Area_Type,
    VEHICLE_COUNT,
    TOTAL_UNITS
  )
```

## Exploratory Data Analysis

### Seasonal Crash Patterns

The tables display four tables showcasing the crash patterns based on road conditions and area type that the crash occurred in.

The dominant pattern across all four seasons is that most crashes occurred on `Dry` roads, with `Property Damage`as the most common severity outcome. The most common collision type was `Hit Fixed Object` signaling that drivers often hit things that weren't moving. The amount of car crashes that occurred on dry roads is unexpected considering how in Winter and Spring the dominant weather types are typically snow (winter) and rain(spring). Summer and Fall's data makes sense when comparing to Winter and Spring due to summer and fall being dryer seasons allowing for more incidents on road. This finding suggests that seasonal factors may not be main contributor to car crashes in Centre County.

```{r}
#| label: seasonal-crash-tables
#| echo: false
#| message: false
#| warning: false
#| fig-alt: Four tables showing crash patterns by road condition, area type,  top collision type, and top severity level for each season in Centre  County, Pennsylvania, 2023-2024.
#| results: asis
# Style Guide: Google R Style Guide
# Primary Author: Akil Creswell
# Reviewed by: Jooan Choi


# Step 1: Define season from month ----

winter <- c("December", "January", "February")
spring <- c("March", "April", "May")
summer <- c("June","July","August")
fall <- c("September", "October", "November")

crashes_seasonal <- crashClean |>
  mutate(
    Season = case_when(
      Month %in% winter ~ "Winter",
      Month %in% spring ~ "Spring",
      Month %in% summer ~ "Summer",
      Month %in% fall ~ "Fall"
    )
  )

# Step 2: Build summary per season/road/area ----
seasonal_summary <- crashes_seasonal |>
  group_by(Season, Road_Condition, Area_Type) |>
  summarise(
    Top_Collision = unique(Collision_Type)[which.max(tabulate(match(
      Collision_Type, unique(Collision_Type)
    )))],
    Top_Severity = unique(Severity_Level)[which.max(tabulate(match(
      Severity_Level, unique(Severity_Level)
    )))],
    Crashes = n(),
    .groups = "drop"
  ) |>
  filter(Crashes >= 5)

# Step 3: Render one table per season ----
seasons <- c("Winter", "Spring", "Summer", "Fall")

cat("\\captionsetup{labelformat=empty}\n")

for (s in seasons) {
  table_output <- seasonal_summary |>
    filter(Season == s) |>
    select(
      Road_Condition, Area_Type, Top_Collision,
      Top_Severity, Crashes
    ) |>
    arrange(Road_Condition, Area_Type) |>
    kable(
      format = "latex",
      caption = paste(
        s, ": Crash patterns by Road Condition",
        "and Area Type (Centre County, 2023-2024)"
      ),
      col.names = c(
        "Road Condition", "Area Type",
        "Top Collision", "Top Severity", "Crashes"
      ),
      align = c("l", "l", "l", "l", "r"),
      booktabs = TRUE
    ) |>
    kable_styling(
  font_size = 11,
  latex_options = c("HOLD_position", "bordered")) |>
  print(table_output)
}

```

{{< pagebreak >}}

### **Interesting Insight**

The plot displays frequency of collision types on each day of the week.

Drivers hitting a fixed object(non moving) was the most common collision type on every day of the week. This specific collision type happened on mostly Fridays and Saturdays. Crashes that were categorized as `Hit Fixed Object` accounted for 722 total incidents in 2023-2024 and was the dominant outcome. This was the highest collision count(by type) of the ten in the datasets. Friday and Saturday combined for 225 of those crashes, showing an end-of-week behavioral pattern. This data aligns with the finding referenced in the article from @romanow2025 that discussed how Friday is Pennsylvania's most common crash day. Some of the causes discussed were heavier traffic, drunk/drugged driving, and Increased fatigue which all seem to be major causes for car crashes.

```{r}
#| label: insight-chart
#| echo: false
#| message: false
#| warning: false
#| fig-alt: "Line chart showing the frequency of each collision type across days of the week in Centre County, Pennsylvania for 2023 and 2024. Hit Fixed Object is the highest line on every day, peaking on Friday and Saturday. Rear-end and Angle collisions are the next most frequent. All other collision types remain low and relatively flat across the week."
#| fig-width: 8
#| fig-height: 7
# Style Guide: Google R Style Guide
# Primary Author: Akil Creswell
# Reviewed by: Jooan Choi

collision_labels <- c(
  "0" = "Non-collision",
  "1" = "Rear-end",
  "2" = "Head-on",
  "3" = "rear-to-rear",
  "4" = "Angle",
  "5" = "Sideswipe",
  "6" = "Sideswipe (Opposite)",
  "7" = "Hit Fixed Object",
  "8" = "Hit Non-Motorist",
  "98" = "Other"
)

day_labels <- c(
  "1" = "Sun",
  "2" = "Mon",
  "3" = "Tue",
  "4" = "Wed",
  "5" = "Thu",
  "6" = "Fri",
  "7" = "Sat"
)


crash_summary <- crashClean |>
  mutate(
    collision_label = case_when(
      Collision_Type == "Non-collision" ~ "Non-collision",
      Collision_Type == "Rear-end" ~ "Rear-end",
      Collision_Type == "Head-on" ~ "Head-on",
      Collision_Type == "Backing" ~ "Backing",
      Collision_Type == "Angle" ~ "Angle",
      Collision_Type == "Sideswipe same direction" ~ "Sideswipe",
      Collision_Type == "Sideswipe opposite direction" ~ "Sideswipe (Opposite)",
      Collision_Type == "Hit fixed object" ~ "Hit Fixed Object",
      Collision_Type == "Hit non-motorist" ~ "Hit Non-Motorist",
      .default = "Other"
    ),
    day_label = factor(
      Day,
      levels = c(
        "Sunday", "Monday", "Tuesday", "Wednesday",
        "Thursday", "Friday", "Saturday"
      )
    )
  ) |>
  filter(!is.na(day_label), !is.na(collision_label)) |>
  group_by(day_label, collision_label) |>
  summarise(count = n(), .groups = "drop")

ggplot(crash_summary, aes(
  x = day_label,
  y = count,
  color = collision_label,
  group = collision_label
)) +
  scale_y_continuous(breaks = seq(0,120,by = 10)) +
  geom_line(na.rm = TRUE) +
  geom_point(aes(shape = collision_label), size = 3, na.rm = TRUE) +
  labs(
    title = "Collision Type Frequency by Day of Week",
    subtitle = "Centre County, Pennsylvania | 2023-2024",
    x = "Day of Week",
    y = "Number of Crashes",
    color = "Collision Type",
    caption = "Source: PennDOT Crash Data",
    shape = "Collision Type"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.key.size = unit(0.4, "cm"),
    legend.text = element_text(size = 7),
    legend.justification = "top",
    legend.margin = margin(t = 0, unit = "pt"),
    plot.title = element_text(size = 15, face = "bold",margin = margin(b = 20, unit = "pt")),
    plot.subtitle = element_text(size = 12, margin = margin(b = 10 ,unit = "pt")),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5),
    axis.title.x = element_text(margin = margin(t = 15, unit = "pt")),
    axis.title.y = element_text(margin = margin(r = 15, unit = "pt")))
```

{{< pagebreak >}}

### Road Condition *x* Severity Heatmap

The heatmap displays crash counts across road conditions, segmented by severity group.

The dominant road condition was `Dry` and most of those crashes were categorized as less severe. This is expected as most likely drivers would travel under dry conditions, naturally producing higher crash volume. It was also expected that the second highest would be `Wet` because that condition is most likely to still be driven since it isn't considered as dangerous as conditions like `Snow` or `Ice or Frost`*.* An unexpected anomaly was that overall there was only 100 car crashes categorized as `More Severe`*.* The lack of recorded crashes with major severity suggests that Road Condition is not as substantial of a factor when it comes to the worst crashes in Centre County.

```{r}
#| label: road-severity-heatmap
#| echo: false
#| message: false
#| warning: false
#| fig-alt: "Heatmap showing crash counts by road condition and severity
#|   group in Centre County, Pennsylvania, 2023-2024. Dry roads dominate
#|   with 1,350 less severe and 79 more severe crashes. Wet roads are
#|   second with 352 less severe and 11 more severe crashes. Mud, dirt,
#|   or gravel roads show the highest proportion of more severe crashes
#|   relative to total volume with 6 more severe out of 16 total."
#| fig-width: 8
#| fig-height: 7
# Style Guide: Google R Style Guide
# Primary Author: Akil Creswell
# Reviewed by: Jooan Choi

# Step 1: Filter and count ----
heatmap_data <- crashClean |>
  filter(
    !is.na(Road_Condition),
    Severity_Group != "Unknown"
  ) |>
  group_by(Road_Condition, Severity_Group) |>
  summarise(Crashes = n(), .groups = "drop")

# Step 2: Order road conditions by total volume ----
road_order <- heatmap_data |>
  group_by(Road_Condition) |>
  summarise(Total = sum(Crashes), .groups = "drop") |>
  arrange(desc(Total)) |>
  pull(Road_Condition)

# Step 3: Plot heatmap ----
ggplot(
  heatmap_data,
  aes(
    x = Severity_Group,
    y = factor(Road_Condition, levels = rev(road_order)),
    fill = Crashes
  )
) +
  geom_tile(color = "#1e407c", linewidth = 0.5) +
  geom_text(
    aes(label = Crashes),
    size = 3
  ) +
  scale_fill_gradient(
    low = "white",
    high = "#96BEE6",
    name = "Crash Count"
  ) +
  labs(
    title = "Crash Severity by Road Condition",
    subtitle = "Centre County, Pennsylvania | 2023-2024",
    x = "Severity Group",
    y = "Road Condition",
    caption = "Source: PennDOT Crash Data"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_text(size = 10),
    axis.title.x = element_text(margin = margin(t = 15,b= 30, unit = "pt")),
    axis.title.y = element_text(margin = margin(r = 15, unit = "pt")),
    plot.title = element_text(size = 15, face = "bold"),
    plot.subtitle = element_text(size = 12),
  )

```

{{< pagebreak >}}

## Summary

This analysis examined 2,097 recorded crashes in Centre County, Pennsylvania from 2023-2024. Through exploratory data analysis and visual summarization, we identified patterns in collision type and road conditions across varying levels of crash severity.

Three patterns emerged consistently across the data. First, Friday and Saturday were the most common days for crashes to occur, suggesting an end-of-week behavioral pattern. The second pattern was discovered while assessing seasonal crash patterns by road condition and area type. The most common road condition throughout all four seasons(winter, spring, summer, fall) was `Dry` with the most prominent severity level being `Property Damage Only`*.* The third pattern was found while assessing the road conditions by crash severity grouping. `Dry` was still the common road condition amongst the crashes and the data showed there was only 100 of 1,961 classified crashes there categorized as `More Severe`*.*

Put together these findings suggest that Road Conditions do not heavily factor into car crash frequency, type, and severity level. In the future these data could be combined with other data to enable a more thorough investigation into the causes of crashes in Centre County such as driver information, specific time of day, and road type(highway, interstate, residential).

{{< pagebreak >}}

## Author Contributions

This project followed the CRediT Contributor Role Taxonomy to document individual contributions.

**Akil** contributed to conceptualization, formal analysis (seasonal crash tables, heatmap), visualization (insight chart), (heatmap), and writing of the introduction, data, EDA, and summary sections.

**Jooan** contributed to data curation (wrangling and cleaning of both CSV files), formal analysis (overview summary table, monthly crash bar chart), and visualization.

**Inayat** contributed to

{{< pagebreak >}}

## **References**

::: {#refs}
:::

{{< pagebreak >}}

## **Code Appendix**

```{r ref.label=knitr::all_labels()}
#| echo: true
#| eval: false
```
